Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
4.1 wip

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. The R internal types are translated to the type name of the proposed
type system. The next bit of information output for each value is its
\emph{class}.  The class is an optional list of names, it may be absent,
and, in some cases, it may be implicit (i.e. the interpreter blesses some
values with the \k{matrix} and \k{array} classes even when they have no
attribute). Finally, for each value, the tool outputs the list of
\emph{attribute names} (other than \k{class}).

@@

\begin{itemize}
\item For vectors, the number of dimensions and their names are output;
\item For lists, a recursive traversal collects element types and
  the name attributes of elements.  
  TODO: HOW DO WE OUTPUT HETEROGENEOUS NAMES???
\item For promises, we carefully dig into the promise value {\it without
  forcing it}, and collect information about the innermost value. If the
  innermost value is the unbound value, then the promise is an as-of-yet
  unevaluated expression. In this case, we do our best to guess the value of
  the unevaluated expression: if the expression is not a language
  expression, we collect information about it, otherwise we can say nothing
  without evaluating it and ascribe it type ``any''. 
\end{itemize}

For vectors (and lists), we additionally iterate through the elements,
checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
part of a function's varargs (denoted \ldots in function definitions) are
tagged as such and ignored, as varargs can contain anything (and their type
will be equivalent to and any-typed list).

@@

%
%
\subsubsection{Challenges}
\label{subsec:typetracer-challenges}

We grappled with a number of corner cases related to R's non-strict semantics.

\begin{itemize}

\item First, we wanted to ensure that \typetracer captured types for
  unevaluated arguments.  Unfortunately, not every argument of a function is
  necessarily used during every function execution, meaning that some
  arguments may remain unevaluated (as they are never forced), and there are
  a few reasons for this.

@@

\item Second, we wanted to account for formal arguments (arguments which are
  named in the function definition) which received no values when the
  function was called. This represents cases where a function was not called
  with all of its arguments, and no default value was specified for missing
  arguments.  We record a ``missing'' type for such argument.

@@

\item Third, we had to contend with R's backend sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  
\end{itemize}

Having outlined these challenges, we will now explore the implementation of \typetracer in more detail.

%
%
\subsubsection{Select Details}

As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
is a very efficient instrumentation framework for R, which is built on a
modified version of the R runtime.

We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
\texttt{special\_entry},\texttt{special\_exit},
\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
callbacks fire when closures, builtin, and special functions are entered and
exited, and before and after the forcing of promises.  The function-related
callbacks are used mainly for bookkeeping: the analysis is notified that a
closure, builtin, or special has been entered/exited by pushing/popping the
call onto a stack.  The calls themselves store a {\it call trace} object,
where we store the type information that we collect, populated with initial
guesses of argument types (as discussed in
Section~\ref{subsec:typetracer-challenges}).  Also, recall that R can
perform single or multiple dispatch on function arguments depending on their
class.  We collect such dispatch information the {\tt \_entry} variants.

The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
As arguments to functions in R are wrapped in promises, we delay our
reflection to promise force time.  In R, promises are essentially objects
with value and expression slots. \AT{Put this as part of R-dyntrace
  background:} When a promise is created for a particular expression, that
expression is merely put into the expression slot of a promise object, and a
special {\it unbound} value is loaded into the value slot (unless some
default value is specified, wherein it is loaded into the value slot---this
is the case with default argument values).  When the promise is forced, the
expression is evaluated in the forcing context, and the value is stored in
the value slot of the promise.  So to get the type of a function argument,
we check the value slot of the argument promise---recursively if the value
itself is a promise.

To construct a type for each argument, we make use of R's C FFI and use
low-level machinery to collect type tags and attributes from the R runtime.
The types that we provide to users are constructed during post-processing,
and rely on the detailed information made available by these low-level
reflection mechanisms.

@@

\subsection{Is our simple type language expressive enough to type R code?} Functions in
R can support arguments of multiple types by reflecting on the type to
perform appropriate conversions and computations. On the one hand, this form
of ad-hoc polymorphism makes R functions expressive; on the other hand, it
requires support for polymorphism in the typing scheme, making the type
system complex.  Our system lacks explicit support for
polymorphism. Instead, it supports union, any and ... types. A union type is
inferred when values of multiple types are observed at a function parameter
and return position, any type is inferred when an argument is missing or
unused, and ... is inferred for vararg parameters. Furthermore, list element
types can be inferred as any if they have more than 5 different types. Values with many different types can result in
a large number of alternatives in the type inferred for that position,
making it hard to read and unwieldy for end-users. To evaluate this
''expressivity gap`' between our type inference scheme and R functions, we
look at the number of different types in the parameter and return types of
the XXX functions in our corpus.

\begin{figure}[!h]  \centering
  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
  \caption{Types inferred for each position}\label{fig:unionfreq}
\end{figure}

>>>>

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. The R internal types are translated to names in the proposed type
system. The next bit of information output is the \emph{class} which is an
optional list of names, it may be absent, and, in some cases, it may be
implicit (i.e. the interpreter blesses some values with the \k{matrix} and
\k{array} classes even when they have no attributes).

@@

\begin{itemize}
\item For vectors, the number of dimensions and the present of \NA values.
\item For lists, a recursive traversal collects element types.
\item For promises, we attempt to establish the type of its result or output
  \any.
\end{itemize}

\noindent
To obtain raw types, we make use of R's C FFI and use low-level machinery to
collect type tags and attributes from the R runtime.  The types that we
provide to users are constructed during post-processing, and rely on the
detailed information made available by these low-level reflection
mechanisms.

\paragraph{Varargs}
Arguments that are part of a function's varargs (denoted \VARGS) are
ignored, as type \any is output for the entire varag construct.

\paragraph{Promises}

@@

First, we wanted to ensure that \typetracer captured types for unevaluated
arguments.  Unfortunately, not every argument of a function is necessarily
used during every function execution, meaning that some arguments may remain
unevaluated (as they are never forced), and there are a few reasons for
this.

@@

\paragraph{Missing arguments}
Second, we wanted to account for formal arguments (arguments which are named
in the function definition) which received no values when the function was
called. This represents cases where a function was not called with all of
its arguments, and no default value was specified for missing arguments.  We
record a ``missing'' type for such argument.

@@

\paragraph{Non-local returns}
Third, we had to contend with R's backend sometimes performing longjumps,
which discards current R calls from the call stack.  In order to ensure that
call traces are not lost when a longjump occurs, \typetracer intercepts the
unwinding process initiated by the longjump and mimics the functionality of
the functions being exited.  The R-dyntrace framework supplies the return
value during the unwinding process (if there is indeed such a value), so we
can capture it, analyze it, and include it in the call trace.  In certain
situations, however, no return value can be obtained (e.g. a longjump
occurred as part of some error handling), and then we simply record a
``jumped'' return value, which we can deal with in post-processing.

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},\k{special\_exit},
\k{promise\_force\_entry}, and \k{promise\_force\_exit}.  The
function-related callbacks are used mainly for bookkeeping: the analysis is
notified that a construct has been entered by pushing the call onto a stack.
The calls themselves store a trace object, it is that object that holds the
type information. As R can perform single or multiple dispatch on function
arguments depending on their class, the relevant information is kept by the
{\tt \_entry} variants.

The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
As arguments to functions in R are wrapped in promises, we delay our
reflection to promise force time.  In R, promises are essentially objects
with value and expression slots.

When a promise is created for a particular expression, that expression is
merely put into the expression slot of a promise object, and a special {\it
  unbound} value is loaded into the value slot (unless some default value is
specified, wherein it is loaded into the value slot---this is the case with
default argument values).  When the promise is forced, the expression is
evaluated in the forcing context, and the value is stored in the value slot
of the promise.  So to get the type of a function argument, we check the
value slot of the argument promise---recursively if the value itself is a
promise.

@@

\subsection{Is our simple type language expressive enough to type R code?} Functions in
R can support arguments of multiple types by reflecting on the type to
perform appropriate conversions and computations. On the one hand, this form
of ad-hoc polymorphism makes R functions expressive; on the other hand, it
requires support for polymorphism in the typing scheme, making the type
system complex.  Our system lacks explicit support for
polymorphism. Instead, it supports union and any types. A union type is
inferred when values of multiple types are observed at a function parameter
and return position. any type is inferred when an argument is missing or
unused. Furthermore, list element types can be inferred as any if they have
more than 5 different types. Values with many different types can result in
a large number of alternatives in the type inferred for that position,
making it hard to read and unwieldy for end-users. To evaluate this
''expressivity gap`' between our type inference scheme and R functions, we
look at the number of different types in the parameter and return types of
the XXX functions in our corpus.
